---
title: purrrplus
description: "An R package that expands the functionality of purrr."
author: ''
date: '2018-06-10'
slug: purrrplus
tags:
  - R
  - Tidyverse
  - Packages
comments: no
showcomments: yes
showpagemeta: yes
---

[**See the package website**](https://github.com/stenhaug/purrrplus)

### Background

In my work, I often need to run a function in R many times (for example in running a simulation). If any of these runs throw an error, all progress is lost. The tidyverse package purrrr provides functionality like *safely* which helps but makes subsequent analysis tricky. As a remedy, I wrote the package purrrplus which allows for running a function safely with easy analysis of errors and results. 

### Example

```{r message=FALSE}
library(purrrplus)
library(tidyverse) # most useful with the tidyverse
```

Imagine you have a function (which returns a named list or a named vector and might throw an error):

```{r}
calculate_if_positive <- function(a, b){
  if(a < 0 & b < 0) {stop("Both numbers are negative.")}
  else if(a < 0) {stop("Just the first number is negative")}
  else if(b < 0) {stop("Just the second number is negative")}
  
  list(add = a + b,
       subtract = a - b,
       multiply = a * b,
       divide = a / b)
}
```

And you want to apply this function to each row of a data frame (which might contain irrelevant variables):

```{r}
(numbers <- data_frame(a = c(-1, 0, 1, 2),
                      b = c(2, 1, 0, -1),
                      irrelevant = c("minneapolis", "st_paul", "minneapolis", "st_paul")))
```

pmap_safely adds an error and a result column (which come from applying the function) to the inputted data frame.

```{r}
(output <- pmap_safely(numbers, calculate_if_positive))
```

get_errors allows for quick analysis of errors:

```{r}
get_errors(output)
```

get_results filters out rows with errors and unnests results such that each item in the list that the function returns has its own column:

```{r}
get_results(output)
```

